---
title: "Preliminary analysis of the available data for BC Butterflies and Dragonflies with a comparison of the data available online from collections and the data available from iNaturalist"
author: "Nathan G. Earley"
date: "2026-01-173"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rinat)
library(rgbif)
library(sf)
```

## Intro

We're trying to figure out where we should do the Southern Interior of BC's Dragonfly Count and Butterfly Count. 

I'm not uploading the data to GitHub because they're fairly big files... let me know if you want them and I'll share.

## Candidate locations

There are 3 Christmas Bird count locations that seem like they would be the best fit for hosting the Ode and Butt Counts. Apex Mountain, Oliver-Osoyoos, and Vaseux Lake.

First we'll get those circles.

For the Oliver-Osoyoos, and Vaseux Lake CBC circles we can use Dave Bell's CBC circles which I downloaded from his "[My Google Maps](https://www.google.com/maps/d/viewer?mid=1D49EyLXIFoIBQYS5rE6Fi9YTenb6H46A&ll=49.715363640775124%2C-119.40967101463359&z=8)" and edited in OkanaganCBCcircles.R in this .RProj directory.

```{r ReadCircles, echo = TRUE}
OliverOsoyoosCBC <- sf::st_read("data/NathanCBCkmls/OliverOsoyoosCBC_Circle.kml",
                                layer = "OliverOsoyoosCBC_Circle",
                                quiet = TRUE)
VaseuxLakeCBC <- sf::st_read("data/NathanCBCkmls/VaseuxLakeCBC_Circle.kml", 
                             layer = "VaseuxLakeCBC_Circle",
                             quiet = TRUE)

## Reproject to EPSG:4326
OliverOsoyoosCBC_wgs84 <- sf::st_transform(OliverOsoyoosCBC, 4326)
VaseuxLakeCBC_wgs84 <- sf::st_transform(VaseuxLakeCBC, 4326)
```

Dave does not have a circle for apex so I'm approximating the centroid of the circles on the Birds Canada CBC map and rolling with that.

```{r MakeApexCircle, echo = TRUE}
## Set the centroid (Note: this is approximate)
ApexCentroid <- sf::st_sfc(
  sf::st_point(c(-119.911581, 49.308790)),  # lon, lat
  crs = 4326)

## Project to EPSG:3005 = BC Albers
ApexCentroid_proj <- sf::st_transform(ApexCentroid, 3005)

## Make the circle (12km radius)
ApexCBCcircle <- sf::st_buffer(ApexCentroid_proj, dist = 12000)

## Reproject to EPSG:4326
ApexCBCcircle_wgs84 <- sf::st_transform(ApexCBCcircle, 4326)
```

## Data in Collections

### Odonate data from Gibson et al (2024)

Now that we have the circles lets bring in the digitized Odonata holdings of the BC Entomology Collections from Gibson et al (2024; Can. Entomol. 156(e42): 1–16. doi:10.4039/tce.2024.38). 

```{r NhcOdeDataReadIn, echo = TRUE}
## Read in Gibson data
GibsonDF <- read.csv("data/GibsonData/Gibson2024CanEntSup002.csv")

## Remove rows with NA Lat Long data
GibsonDF <- GibsonDF %>% 
  dplyr::filter(!is.na(Longitude), !is.na(Latitude))

## Restrict the datset to just BC records
GibsonBC <- GibsonDF %>% 
  dplyr::filter(Prov_State == "British Columbia")

## Finally we need to make sure that the Gibson data are in the right projection
GibsonBC_sf <- GibsonBC %>% 
  sf::st_as_sf(coords = c("Longitude", "Latitude"),
               crs = 4326,
               remove = FALSE)
```

Now that we've got that sorted we can subset the data in the Gibson dataset to those records within these circles.

```{r NhcOdeCut2Circles, echo = TRUE, warning = FALSE}
## Filter the data to the circles
ApexOdeRecords <- GibsonBC_sf %>% 
  dplyr::filter(sf::st_intersects(., ApexCBCcircle_wgs84, sparse = FALSE)[, 1])
OliverOdeRecords <- GibsonBC_sf %>% 
  dplyr::filter(sf::st_intersects(., OliverOsoyoosCBC_wgs84, sparse = FALSE)[, 1])
VaseuxOdeRecords <- GibsonBC_sf %>% 
  dplyr::filter(sf::st_intersects(., VaseuxLakeCBC_wgs84, sparse = FALSE)[, 1])
```

Now we can do some quick summary stats to see how much data is in each dataset

```{r NhcOdeSummStats, echo = TRUE}
## How many specimens in each circle
nrow(ApexOdeRecords)
nrow(OliverOdeRecords)
nrow(VaseuxOdeRecords)

## How many species (ItemName) in each circle
length(unique(ApexOdeRecords$ItemName))
length(unique(OliverOdeRecords$ItemName))
length(unique(VaseuxOdeRecords$ItemName))

## Now lets see how many of each species
ApexOdeRecSppCounts <- ApexOdeRecords %>%
  dplyr::group_by(ItemName) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count)) 
OliverOdeRecSppCounts <- OliverOdeRecords %>%
  dplyr::group_by(ItemName) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
VaseuxOdeRecSppCounts <- VaseuxOdeRecords %>%
  dplyr::group_by(ItemName) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))

## What are the top 5 spp and how many records do they have?
head(sf::st_drop_geometry(ApexOdeRecSppCounts))
head(sf::st_drop_geometry(OliverOdeRecSppCounts))
head(sf::st_drop_geometry(VaseuxOdeRecSppCounts))
```

How many species have 1 specimen, 2 specimens, etc?

```{r NhcOdeCircleSppHists, echo = TRUE}
ggplot(ApexOdeRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode records per Species",
    y = "Number of Species",
    title = "Distribution of Ode Records per Species (Apex)"
  ) +
  theme_minimal()

ggplot(OliverOdeRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode records per Species",
    y = "Number of Species",
    title = "Distribution of Ode Records per Species (Oliver)"
  ) +
  theme_minimal()

ggplot(VaseuxOdeRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode records per Species",
    y = "Number of Species",
    title = "Distribution of Ode Records per Species (Vaseux)"
  ) +
  theme_minimal()
```

Now we know what's going on in the BC ode collections according to Gibson et al (2024) for these circles.

### Butterfly data from GBIF

Now that we have the circles lets bring in the digitized butterfly data available on GBIF and downloaded on Jan 15, 2026.

```{r NhcButtDataReadIn, echo = TRUE}
GbifHesp <- readr::read_tsv("data/GbifData/Hesperiidae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifLyca <- readr::read_tsv("data/GbifData/Lycaenidae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifNymp <- readr::read_tsv("data/GbifData/Nymphalidae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifPapi <- readr::read_tsv("data/GbifData/Papilionidae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifPier <- readr::read_tsv("data/GbifData/Pieridae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifRiod <- readr::read_tsv("data/GbifData/Riodinidae/occurrence.txt",
                            na = c("", "NA"),
                            show_col_types = FALSE) %>% 
  dplyr::mutate(decimalLatitude  = as.numeric(decimalLatitude),
                decimalLongitude = as.numeric(decimalLongitude),
                year = as.integer(year),
                georeferencedDate= as.character(georeferencedDate))

GbifButt <- dplyr::bind_rows(GbifHesp, GbifLyca, GbifNymp, 
                             GbifPapi, GbifPier, GbifRiod)

## Remove obs that aren't IDed to spp
GbifButt_clean <- GbifButt %>% 
  dplyr::filter(!is.na(acceptedScientificName) & species != "") %>%
  dplyr::filter(!str_detect(acceptedScientificName, 
                            regex("complex", ignore_case = TRUE))) %>% 
  dplyr::mutate(
    acceptedScientificName = dplyr::if_else(
      stringr::str_starts(acceptedScientificName, "BOLD"),
      acceptedScientificName,
      acceptedScientificName %>%
        # remove "subsp"
        str_remove_all("\\bsubsp\\b") %>% 
        # remove "var." and "var"
        str_remove_all("\\bvar\\.\\b") %>% 
        str_remove_all("\\bvar\\b") %>% 
        # keep up to first 3 words, treating hyphenated words as one
        str_replace("^((?:\\b[\\w-]+\\b\\s+){0,2}\\b[\\w-]+\\b).*", "\\1") %>% 
        # remove text in parentheses
        str_remove_all("\\s*\\([^)]*\\)") %>%
        # remove numbers
        str_remove_all("\\d+") %>%
        # remove word before comma + comma
        str_remove_all("\\b\\w+\\s*,") %>%
        # remove word or letter before period
        str_remove_all("\\b\\w+\\.") %>%
        # remove trailing quotes
        str_remove_all('"$') %>%
        # remove capitalized words after the second word (genus + species) but keep hyphenated species
        str_replace("^([A-Z][a-z]+\\s+[a-z-]+(?:\\s+[a-z-]+)?).*", "\\1") %>% 
        # remove '||''||' and anything after
        str_remove("\\|\\|''\\|\\|.*$") %>%
        # remove "&"s
        str_remove_all("&") %>%
        # remove any remaining [] brackets
        str_remove_all("\\[|\\]") %>%
        # remove second Parnassius and after
        str_replace_all("(?<=\\bParnassius\\b).*\\bParnassius\\b.*$", "") %>%
        # remove second Papilio and after
        str_replace_all("(?<=\\bPapilio\\b).*\\bPapilio\\b.*$", "") %>%
        # clean leftover whitespace
        str_squish()
    )
  ) %>%
  dplyr::filter(str_count(acceptedScientificName, "\\S+") > 1)

sort(unique(GbifButt_clean$acceptedScientificName))
```

The GBIF download DOI is `meta$doi`.

Now that we've got that sorted we can subset the data in the GBIF dataset to those records within these circles.

```{r NhcButtCut2Circles, echo = TRUE, warning = FALSE}
## Filter the data to the circles
ApexButtRecords <- GbifButtData %>% 
  dplyr::filter(sf::st_intersects(., ApexCBCcircle_wgs84, sparse = FALSE)[, 1])
OliverButtRecords <- GbifButtData %>% 
  dplyr::filter(sf::st_intersects(., OliverOsoyoosCBC_wgs84, sparse = FALSE)[, 1])
VaseuxButtRecords <- GbifButtData %>% 
  dplyr::filter(sf::st_intersects(., VaseuxLakeCBC_wgs84, sparse = FALSE)[, 1])
```

Now we can do some quick summary stats to see how much data is in each dataset

```{r NhcButtSummStats, echo = TRUE}
## How many specimens in each circle
nrow(ApexButtRecords)
nrow(OliverButtRecords)
nrow(VaseuxButtRecords)

## How many species in each circle
length(unique(ApexButtRecords$species))
length(unique(OliverButtRecords$species))
length(unique(VaseuxButtRecords$species))

## Now lets see how many of each species
ApexButtRecSppCounts <- ApexButtRecords %>%
  dplyr::group_by(species) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count)) 
OliverButtRecSppCounts <- OliverButtRecords %>%
  dplyr::group_by(species) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
VaseuxButtRecSppCounts <- VaseuxButtRecords %>%
  dplyr::group_by(species) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))

## What are the top 5 spp and how many records do they have?
head(sf::st_drop_geometry(ApexButtRecSppCounts))
head(sf::st_drop_geometry(OliverButtRecSppCounts))
head(sf::st_drop_geometry(VaseuxButtRecSppCounts))
```

How many species have 1 specimen, 2 specimens, etc?

```{r NhcButtCircleSppHists, echo = TRUE}
ggplot(ApexButtRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt records per Species",
    y = "Number of Species",
    title = "Distribution of Butt Records per Species (Apex)"
  ) +
  theme_minimal()

ggplot(OliverButtRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt records per Species",
    y = "Number of Species",
    title = "Distribution of Butt Records per Species (Oliver)"
  ) +
  theme_minimal()

ggplot(VaseuxButtRecSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt records per Species",
    y = "Number of Species",
    title = "Distribution of Butt Records per Species (Vaseux)"
  ) +
  theme_minimal()
```

Now we know what's going on in the BC Butt collections according to GBIF for these circles.

## Data on iNaturalist

Lets look at iNat to see how many records there are in the iNat dataset for Odonata and Papilionoidea. This dataset is much younger but will likely have more recent data. The data we'll use were downloaded from iNaturalist through the Export function on 2026-01-13. Note, some of these records are obscured... we'll start by not removing them but keep that in mind.

### Odonate iNat data

Now we'll read in the iNat data. 

```{r iNatOdeDataReadIn, echo = TRUE}
## Read in the data
iNatOdeDF <- read.csv("data/iNatData/iNatOdesBC2026JAN13.csv")

## Keep only research grade obs
iNatOdeRG <- iNatOdeDF %>% 
  dplyr::filter(quality_grade == "research")

## Keep only CC0, CC-BY, CC-BY-NC records so you don't go to jail for 1B years 
iNatOdeRG_CC <- iNatOdeRG %>% 
  dplyr::filter(license %in% c("CC0", "CC-BY", "CC-BY-NC"))

## Remove read.csv2()## Remove all data with big (>5km) or NA accuracy 
iNatOdeRG_CC_acc <- iNatOdeRG_CC %>% 
  dplyr::filter(positional_accuracy < 5000)

## Change the Lat and long columns to Lat and Long
iNatOdeRG_CC_acc <- iNatOdeRG_CC_acc %>% 
  dplyr::rename("Longitude" = "longitude") %>% 
  dplyr::rename("Latitude" = "latitude")

## Finally we need to make sure that the iNat data are in the right projection
iNatOde_sf <- iNatOdeRG_CC_acc %>% 
  sf::st_as_sf(coords = c("Longitude", "Latitude"),
               crs = 4326,
               remove = FALSE)
```

Now that we have the iNat data read in and cleaned lets filter to the same circles as before.

```{r iNatOdeCut2Circles, echo = TRUE, warning = FALSE}
## Filter the data to the circles
ApexOdeObs <- iNatOde_sf %>% 
  dplyr::filter(sf::st_intersects(., ApexCBCcircle_wgs84, sparse = FALSE)[, 1])
OliverOdeObs <- iNatOde_sf %>% 
  dplyr::filter(sf::st_intersects(., OliverOsoyoosCBC_wgs84, sparse = FALSE)[, 1])
VaseuxOdeObs <- iNatOde_sf %>% 
  dplyr::filter(sf::st_intersects(., VaseuxLakeCBC_wgs84, sparse = FALSE)[, 1])
```

Now we can do some quick summary stats to see how much data is in each dataset

```{r iNatOdeSummStats, echo = TRUE}
## How many specimens in each circle
nrow(ApexOdeObs)
nrow(OliverOdeObs)
nrow(VaseuxOdeObs)

## How many species (scientific_name) in each circle
length(unique(ApexOdeObs$scientific_name))
length(unique(OliverOdeObs$scientific_name))
length(unique(VaseuxOdeObs$scientific_name))

## Now lets see how many of each species
ApexOdeObsSppCounts <- ApexOdeObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
OliverOdeObsSppCounts <- OliverOdeObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
VaseuxOdeObsSppCounts <- VaseuxOdeObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))

## What are the top 5 spp and how many records do they have?
head(sf::st_drop_geometry(ApexOdeObsSppCounts))
head(sf::st_drop_geometry(OliverOdeObsSppCounts))
head(sf::st_drop_geometry(VaseuxOdeObsSppCounts))
```

How many species have 1 specimen, 2 specimens, etc?

```{r iNatOdeCircleSppHists, echo = TRUE}
ggplot(ApexOdeObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode observations per Species",
    y = "Number of Species",
    title = "Distribution of Ode Observations per Species (Apex)"
  ) +
  theme_minimal()

ggplot(OliverOdeObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode observations per Species",
    y = "Number of Species",
    title = "Distribution of Ode Observations per Species (Oliver)"
  ) +
  theme_minimal()

ggplot(VaseuxOdeObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Ode observations per Species",
    y = "Number of Species",
    title = "Distribution of Ode Observations per Species (Vaseux)"
  ) +
  theme_minimal()
```

Now that we've done that we can see that the two datasets are (at least) superficially similar for Odonates.

### Butterfly iNat data

Now we'll read in the iNat data. 

```{r iNatButtDataReadIn, echo = TRUE}
## Read in the data
iNatButtDF <- read.csv("data/iNatData/iNatButtsBC2026JAN13.csv")

## Keep only research grade obs
iNatButtRG <- iNatButtDF %>% 
  dplyr::filter(quality_grade == "research")

## Keep only CC0, CC-BY, CC-BY-NC records so you don't go to jail for 1B years 
iNatButtRG_CC <- iNatButtRG %>% 
  dplyr::filter(license %in% c("CC0", "CC-BY", "CC-BY-NC"))

## Remove read.csv2()## Remove all data with big (>5km) or NA accuracy 
iNatButtRG_CC_acc <- iNatButtRG_CC %>% 
  dplyr::filter(positional_accuracy < 5000)

## Change the Lat and long columns to Lat and Long
iNatButtRG_CC_acc <- iNatButtRG_CC_acc %>% 
  dplyr::rename("Longitude" = "longitude") %>% 
  dplyr::rename("Latitude" = "latitude")

## Finally we need to make sure that the iNat data are in the right projection
iNatButt_sf <- iNatButtRG_CC_acc %>% 
  sf::st_as_sf(coords = c("Longitude", "Latitude"),
               crs = 4326,
               remove = FALSE)
```

Now that we have the iNat data read in and cleaned lets filter to the same circles as before.

```{r iNatButtCut2Circles, echo = TRUE, warning = FALSE}
## Filter the data to the circles
ApexButtObs <- iNatButt_sf %>% 
  dplyr::filter(sf::st_intersects(., ApexCBCcircle_wgs84, sparse = FALSE)[, 1])
OliverButtObs <- iNatButt_sf %>% 
  dplyr::filter(sf::st_intersects(., OliverOsoyoosCBC_wgs84, sparse = FALSE)[, 1])
VaseuxButtObs <- iNatButt_sf %>% 
  dplyr::filter(sf::st_intersects(., VaseuxLakeCBC_wgs84, sparse = FALSE)[, 1])
```

Now we can do some quick summary stats to see how much data is in each dataset

```{r iNatButtSummStats, echo = TRUE}
## How many specimens in each circle
nrow(ApexButtObs)
nrow(OliverButtObs)
nrow(VaseuxButtObs)

## How many species (scientific_name) in each circle
length(unique(ApexButtObs$scientific_name))
length(unique(OliverButtObs$scientific_name))
length(unique(VaseuxButtObs$scientific_name))

## Now lets see how many of each species
ApexButtObsSppCounts <- ApexButtObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
OliverButtObsSppCounts <- OliverButtObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))
VaseuxButtObsSppCounts <- VaseuxButtObs %>%
  dplyr::group_by(scientific_name) %>% 
  dplyr::summarise(count = n(), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(count))

## What are the top 5 spp and how many records do they have?
head(sf::st_drop_geometry(ApexButtObsSppCounts))
head(sf::st_drop_geometry(OliverButtObsSppCounts))
head(sf::st_drop_geometry(VaseuxButtObsSppCounts))
```

How many species have 1 specimen, 2 specimens, etc?

```{r iNatButtCircleSppHists, echo = TRUE}
ggplot(ApexButtObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt observations per Species",
    y = "Number of Species",
    title = "Distribution of Butt Observations per Species (Apex)"
  ) +
  theme_minimal()

ggplot(OliverButtObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt observations per Species",
    y = "Number of Species",
    title = "Distribution of Butt Observations per Species (Oliver)"
  ) +
  theme_minimal()

ggplot(VaseuxButtObsSppCounts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Butt observations per Species",
    y = "Number of Species",
    title = "Distribution of Butt Observations per Species (Vaseux)"
  ) +
  theme_minimal()
```

Now that we've done that we can see that the two datasets are (at least) superficially similar for Butterflies.

## Comparative analyses

How does this all look between datasets and taxa?

```{r , echo = TRUE}
## Odonata Museum N Specimens
nrow(ApexOdeRecords)
nrow(OliverOdeRecords)
nrow(VaseuxOdeRecords)

## Papilionoidea Museum N Specimens
nrow(ApexButtRecords)
nrow(OliverButtRecords)
nrow(VaseuxButtRecords)

## Odonata iNat N Observations
nrow(ApexOdeObs)
nrow(OliverOdeObs)
nrow(VaseuxOdeObs)

## Papilionoidea iNat N Observations
nrow(ApexButtObs)
nrow(OliverButtObs)
nrow(VaseuxButtObs)

## Odonata iNat N Species
length(unique(ApexOdeObs$scientific_name))
length(unique(OliverOdeObs$scientific_name))
length(unique(VaseuxOdeObs$scientific_name))

## Papilionoidea iNat N Species
length(unique(ApexButtObs$scientific_name))
length(unique(OliverButtObs$scientific_name))
length(unique(VaseuxButtObs$scientific_name))

## Odonata Museum N Species
length(unique(ApexOdeRecords$ItemName))
length(unique(OliverOdeRecords$ItemName))
length(unique(VaseuxOdeRecords$ItemName))

## Papilionoidea Museum N Species
length(unique(ApexButtRecords$species))
length(unique(OliverButtRecords$species))
length(unique(VaseuxButtRecords$species))

summary_df <- data.frame(
  Site = c("Apex", "Oliver", "Vaseux"),

  ## Odonata — Museum
  Odonata_Museum_Specimens = c(
    nrow(ApexOdeRecords),
    nrow(OliverOdeRecords),
    nrow(VaseuxOdeRecords)
  ),
  Odonata_Museum_Species = c(
    length(unique(ApexOdeRecords$ItemName)),
    length(unique(OliverOdeRecords$ItemName)),
    length(unique(VaseuxOdeRecords$ItemName))
  ),

  ## Odonata — iNaturalist
  Odonata_iNat_Observations = c(
    nrow(ApexOdeObs),
    nrow(OliverOdeObs),
    nrow(VaseuxOdeObs)
  ),
  Odonata_iNat_Species = c(
    length(unique(ApexOdeObs$scientific_name)),
    length(unique(OliverOdeObs$scientific_name)),
    length(unique(VaseuxOdeObs$scientific_name))
  ),

  ## Papilionoidea — Museum
  Papilionoidea_Museum_Specimens = c(
    nrow(ApexButtRecords),
    nrow(OliverButtRecords),
    nrow(VaseuxButtRecords)
  ),
  Papilionoidea_Museum_Species = c(
    length(unique(ApexButtRecords$species)),
    length(unique(OliverButtRecords$species)),
    length(unique(VaseuxButtRecords$species))
  ),

  ## Papilionoidea — iNaturalist
  Papilionoidea_iNat_Observations = c(
    nrow(ApexButtObs),
    nrow(OliverButtObs),
    nrow(VaseuxButtObs)
  ),
  Papilionoidea_iNat_Species = c(
    length(unique(ApexButtObs$scientific_name)),
    length(unique(OliverButtObs$scientific_name)),
    length(unique(VaseuxButtObs$scientific_name))
  ),

  stringsAsFactors = FALSE
)

## Flip the table so it's more readable
summary_df_all <- summary_df |>
  tibble::column_to_rownames("Site") |>
  t() |>
  as.data.frame()

summary_df_spp <- summary_df_all %>%
  tibble::rownames_to_column("metric") %>%
  dplyr::filter(grepl("Species", metric)) %>%
  tibble::column_to_rownames("metric")
  
summary_df_obs <- summary_df_all %>%
  tibble::rownames_to_column("metric") %>%
  dplyr::filter(!grepl("Species", metric)) %>%
  tibble::column_to_rownames("metric")
```

```{r CompTableAll}
knitr::kable(
  summary_df_all,
  caption = "Summary of Odonata and Papilionoidea records and observations by site",
  booktabs = TRUE,
  longtable = FALSE
)
```

```{r CompTableSpp}
knitr::kable(
  summary_df_spp,
  caption = "Summary of Odonata and Papilionoidea species by site",
  booktabs = TRUE,
  longtable = FALSE
)
```

```{r CompTableObs}
knitr::kable(
  summary_df_obs,
  caption = "Summary of Odonata and Papilionoidea records by site",
  booktabs = TRUE,
  longtable = FALSE
)
```





